---
title: "Hearing & Perception"
date: 2026-09-15
order: 10
format:
  revealjs:
    output-file: "10_perception.html"
  html: default
  typst: default    
editor: visual
engine: knitr
echo: false
message: false
warning: false
filters: 
  - typst-render
resources: 
  - audio/*
  - images/*
  - silent_center/silences/*
---

```{r}
source(here::here("_extensions/mcanouil/typst-render/_resources/typst_define.R"))
source(here::here("_defaults.R"))
library(tidynorm)
library(tidyverse)
library(scales)
```

## Hearing vs Perception

Hearing

: Physical and anatomical processes in the ear

Perception

: Sense making of what was heard

## Hearing vs Perception

What you *hear* ≠ what you *percieve*

## Anatomy

Outer Ear

- pinna

- ear canal

- tympanic membrane

## Anatomy

Middle Ear

- [malleus](https://en.wiktionary.org/wiki/malleus#Latin)

- [incus](https://en.wiktionary.org/wiki/incus#Latin)

- [stapes](https://en.wiktionary.org/wiki/stapes#Latin)

## Anatomy

Inner Ear

- Cochlea

## Cochlea

- The shortest possible refractory period of a nerve firing is about 0.5 ms

- That means the highest frequency a nerve could *possibly* send signals at is 2000 Hz

## Cochlea


::::::: {.content-hidden when-format="typst"}
:::::: {layout-ncol="3"}
::: {}
8000 Hz

{{< audio "assets/hello.wav" >}}
:::

::: {}
2000 Hz

{{< audio "assets/hello_4000.wav" >}}
:::

::: {}
1000 Hz

{{< audio "assets/hello_2000.wav" >}}
:::
::::::
::::::: 
## Cochlea (unfurled)

```{typst}
#import "@preview/cetz:0.5.1"
#set text(font: "Noto Sans")
#let freq2mm(freq, A: 165.4, a: 2.1, length: 35, k: 0.88) = {
  calc.log(
    (freq/A) + k,
    base: 10
  ) * (length / a)
}

#let mm2freq(
  mm,
  A: 165.4, 
  a: 2.1, 
  length: 35, 
  k: 0.88
) = {
  (calc.pow(10, (mm / (length / a))) - k) * A
}

#let cols = color.map.turbo
#let ncol = cols.len() - 1

#let lowest = 200
#let highest = 16000
#let frange = highest - lowest

#let linear_f = range(0, ncol).map(
   x => {
     let pct = x/ncol
     (pct * frange) + lowest
   }
)

#let full_mm = range(0,(ncol + 1)).map(x => (x/ncol) * 35)
#let full_hz = full_mm.map(x => mm2freq(x))
#let hz_min = calc.min(..full_hz)
#let hz_max = calc.max(..full_hz)
#let hz_range = hz_max - hz_min
#let hz_pct = full_hz.map(x => (x - hz_min)/hz_range)

#full_mm.insert(0,0)
#full_mm.push(35)

#let pal = hz_pct.map(
  x => {
    let idx = int(calc.round((1-x) * ncol))
    cols.at(idx)
  }
)

#let xend = 22

#let seqs = (
  (
    (0.1,-0.2),
    ((xend - 2)/2,0.3),
    (xend - 2,1)
  ),
  (
    (xend - 2,1),
    ((xend - 1), 2),
    (xend - 2,3)
  ),
  (
    (xend - 2,3),
    ((xend - 2) / 2, 3.6),
    (0.1,3.8)
  )
)

#let seqs2 = (
  (xend - 2,2.3),
  (0,2),
  (xend - 2,1.7)
)

#let fseq = (125, 250, 500, 1000, 2000, 4000, 8000, 16000)
#let mmseq = fseq.map(x => 1-freq2mm(x)/35)
#let xspot = mmseq.map(x => x * (xend - 0.8))

#let n_mm = 7
#let even_pct = range(0,n_mm).map(x => 1 - (x / (n_mm - 1)))
#let even_mm = even_pct.map(x => x * 35)
#let edges = (even_mm.remove(0), even_mm.remove(-1))
#let even_hz = even_mm.map(x => calc.round(mm2freq(x)))

#let sw = 2pt

#cetz.canvas({
  import cetz.draw: *
  circle(
    (0.1,0.9),
    radius: (0.5,2.2/2),
    stroke: sw,
    name: "round-window"
  )
  circle(
    (0.15,2.9),
    radius: (0.4,1.8/2),
    stroke: sw,
    name: "oval-window"
  )
  compound-path({
    for x in seqs {
      bezier-through(..x)
    }
  },
  stroke: sw
)

  compound-path({
    line(
      ..seqs2,
      //close: true
    )    
    bezier-through(
      (xend - 2, 1.7),
      (xend - 1.8, 2),
      (xend - 2, 2.3)
    )
  },
  fill: gradient.linear(..pal.rev()),
  stroke: none
  )

  for s_f in xspot.zip(fseq) {
    circle(
      (s_f.at(0), 2),
      name: str(s_f.at(1)),
      radius: 2pt,
      fill: black
    )
    content(
      (s_f.at(0),2.3),
      [#highlight(fill: white.transparentize(30%))[#s_f.at(1) Hz]],
      anchor: "west",
      angle: 45deg
    )
  }

  content(
    "oval-window.north",
    anchor: "south-west",
    angle: 45deg,
    padding: 0.1,
    [oval window]
  )

  content(
    "round-window.south",
    anchor: "east",
    angle: 45deg,
    padding: 0.3,
    [round window]
  )
  
})

```

## Perception - Nonlinear

```{r}
#| fig-format: svg
tibble(
  Hz = seq(20, 20000, length = 20),
  bark = hz_to_bark(Hz)
) |> 
  ggplot(
    aes(Hz, bark)
  ) +
  geom_point() +
  geom_line() +
  theme(aspect.ratio = 1)
```

## Perception - The Brain

> We do not understand exactly how the brain creates the perception of language.
> (Zsiga, 2024).

## Perception - Speech Variability

```{r}
data("pb52", package = "phonTools")
```

```{r}
pb52 |> 
  mutate(
    id = row_number() 
  ) |> 
  pivot_longer(
    f1:f3
  ) |> 
  mutate(
    formant_n = str_extract(name, "\\d") |> 
      as.numeric(),
    delta_f = value / (formant_n - 0.5)
  ) |> 
  summarise(
    .by = speaker,
    delta_f = mean(delta_f)
  ) |> 
  mutate(
    vtl = 34300/(2*delta_f)
  ) ->
  vtl_ests
```

```{r}
#| fig-width: 5
#| fig-asp: 1
#| fig-align: "center"
#| fig-format: svg
pb52 |> 
  left_join(vtl_ests) |> 
  filter(
    vtl == max(vtl) | vtl == min(vtl),
    repetition == 1
  ) |> 
  mutate(
    ipa = recode_values(
      vowel,
      "{"  ~ "æ",
      "3'" ~ "ɝ",
      "A"  ~ "ɑ",
      "E"  ~ "ɛ",
      "i"  ~ "iː",
      "I"  ~ "ɪ",
      "O"  ~ "ɔ",
      "u"  ~ "uː",
      "U"  ~ "ʊ",
      "V"  ~ "ʌ"
    )
  ) |> 
  ggplot(
    aes(f2, f1, color = as.factor(speaker))
  ) + 
  ggforce::geom_mark_hull(
    concavity = 4,
    aes(fill = as.factor(speaker)),
    alpha = 0.05
  ) +  
  geom_text(
    aes(
      label = ipa
    ),
    family = "Noto Sans"
  ) + 
  ggforce::geom_mark_hull(
    concavity = 4
  ) +
  scale_x_continuous(
    trans = c("log10", "reverse"),
    expand = expansion(mult = 0.2)
  ) +
  scale_y_continuous(
    trans = c("log10", "reverse"),
    expand = expansion(mult = 0.2)
  ) +
  guides(
    color = "none",
    fill = "none"
  ) +
  coord_fixed()
```

## Perception - Speech Variability

```{r}
#| fig-width: 5
#| fig-asp: 0.8
#| fig-align: "center"
#| fig-format: svg

make_hull <- function(data) {
  hull <- chull(x = data$x, y = data$y)
  data.frame(x = data$x[hull], y = data$y[hull])
}

pb52 |> 
  left_join(vtl_ests) |> 
  mutate(
    speaker = fct_reorder(
      factor(speaker), vtl
    )
  ) |> 
  ggplot(
    aes(f2, f1, group = speaker)
  ) + 
  # stat_manual(
  #   geom = "polygon",
  #   fun  = make_hull,
  #   aes(color = vtl),
  #   fill = NA,
  #   alpha = 0.6
  # ) +
  ggforce::geom_mark_hull(
    aes(color = vtl),
    concavity = 5,
    alpha = 0.05,
    alpha = 0.1
  )  +
  scale_x_continuous(
    trans = c("log10", "reverse"),
    expand = expansion(mult = 0.2)
  ) +
  scale_y_continuous(
    trans = c("log10", "reverse"),
    expand = expansion(mult = 0.2)
  ) + 
  scale_color_continuous(
    labels = label_glue("{x} cm")
  ) +
  guides(
    color = guide_colorsteps()
  ) +
  coord_fixed()
```

## Perception - Cue Integration

{{< video https://www.youtube.com/watch?v=aFPtc8BVdJk width="800" height="500" >}}

## Perception - Predictive

```{ojs}
viewof spect2 = {
  const img = html`<img src="./images/silent_${step2}.svg"></img>`
  return img;
}
viewof myAudio2 = {
  const audio = html`<audio controls src="./silent_center/silences/clip_${step2}.wav"></audio>`;
  // Expose the audio element's state or actions reactively
  return audio;
}
viewof step2 = Inputs.range([1, 10], {value: 10, step: 1, label: "step"})
```

## Perception - Categorical (?)

```{ojs}
viewof spect = {
  const img = html`<img src="./images/clip_${step}.svg"></img>`
  return img;
}
viewof myAudio = {
  const audio = html`<audio controls src="./audio/clip_${step}.wav"></audio>`;
  // Expose the audio element's state or actions reactively
  return audio;
}
viewof step = Inputs.range([1, 20], {value: 20, step: 1, label: "step"})

```
