---
title: "_scratch"
format: html
editor: visual
---

---
title: "Sound"
format: revealjs
editor: visual
---

## What is Sound?

```{r}
source(here::here("_defaults.R"))
library(tidyverse)
library(gganimate)
library(emojifont)
library(ggdiagram)
library(scico)
library(scales)
library(geomtextpath)
```

```{r}
make_cycle <- function(df){
  df |> 
    mutate(
      nudge = case_when(
        x <= cycle*2 ~ cos((x - cycle) * 2 * 2 * pi),
        .default = cos((x - cycle) * 2 * 2 * pi) / ((((x/(cycle*2))-1)*5)+1)
      )
    )
}

make_cycle2 <- function(df){
  df |> 
    mutate(
      nudge = cos((x - cycle) * 2 * 2 * pi)
    )
}

make_cycle3 <- function(df){
  df |> 
    mutate(
      nudge = cos((x + cycle) * 2 * 2 * pi)
    )
}

make_sine <- function(df){
  df |> 
    mutate(
      y =  case_when(
        x <= cycle*2 ~ sin((x - cycle) * 2 * 2 * pi),
        .default = sin((x - cycle) * 2 * 2 * pi) / ((((x/(cycle*2))-1)*5)+1)
      )
    )
}

make_sine2 <- function(df){
  df |> 
    mutate(
      y =  sin((x - cycle) * 2 * 2 * pi)
    )
}

make_sine3 <- function(df){
  df |> 
    mutate(
      y =  sin((x + cycle) * 2 * 2 * pi)
    )
}
```

## Not Sound

```{r}
set.seed(2025)
cloud <- tibble(
  y = runif(1000),
  x = runif(1000, min = 0, max = 2)
) |> 
  mutate(rowid = row_number())
 
cloud |> 
  ggplot(
    aes(x, y)
  )+
  geom_point() +
  geom_segment(
    aes(
      y = -0.2,
      yend = -0.2,
      x = 0, 
      xend = 2
    )
  ) +
  coord_fixed()+
  theme_no_x()+
  theme_no_y()  
```

## Sound

```{r}
uniform <- tibble(
  x = seq(0, 2, length = 500)
)

seq(0, 3, length = 300) |> 
  map(
    \(.x) cloud |> mutate(cycle = .x)
  ) |> 
  list_rbind() ->
  sound

seq(0, 3, length = 300) |> 
  map(
    \(.x) uniform |> mutate(cycle = .x)
  ) |> 
  list_rbind() ->
  wave

freq <- 3
nudge_scale <- 0.03
amp <- 0.2

sound |> 
  ggplot(
    aes(x,y, cycle = cycle)
  ) +
  geom_point(
    aes(
      nudge_x = after_stat(nudge) * nudge_scale
    ),
    stat = "manual",
    fun = make_cycle,
    position = position_nudge()
  ) + 
  # geom_line(
  #   data = wave,
  #   stat = "manual",
  #   fun = make_sine,
  #   aes(
  #      y = (after_stat(y) * amp) - 0.2
  #   )
  # ) +
 # geom_text(
 #    data = sound |> filter(rowid == 1),
 #    label = fontawesome("fa-user"),
 #    family='fontawesome-webfont',
 #    color = "red",
 #    aes(
 #      nudge_x = after_stat(nudge) * nudge_scale
 #    ),
 #    stat = "manual",
 #    fun = make_cycle,
 #    position = position_nudge()
 #  ) +
  transition_time(cycle) +
  coord_fixed()+
  ease_aes('linear') +
  labs(title = 'cycle: {sprintf("%.2f", frame_time)}') +
  theme_no_x()+
  theme_no_y()
```

## Sound

```{r}
sound |> 
  ggplot(
    aes(x,y, cycle = cycle)
  ) +
  geom_point(
    aes(
      nudge_x = after_stat(nudge) * nudge_scale
    ),
    stat = "manual",
    fun = make_cycle2,
    position = position_nudge()
  ) + 
  # geom_line(
  #   data = wave,
  #   stat = "manual",
  #   fun = make_sine2,
  #   aes(
  #      y = (after_stat(y) * amp) - 0.2
  #   )
  # ) +
 geom_text(
    data = sound |> filter(rowid == 1),
    label = fontawesome("fa-user"),
    family='fontawesome-webfont',
    color = "red",
    aes(
      nudge_x = after_stat(nudge) * nudge_scale
    ),
    stat = "manual",
    fun = make_cycle2,
    position = position_nudge()
  ) +
  transition_time(cycle) +
  coord_fixed()+
  ease_aes('linear') +
  labs(title = 'cycle: {sprintf("%.2f", frame_time)}') +
  theme_no_x()+
  theme_no_y()
```

Sound is not the air molecules from my lungs hitting your ears.

## Sound

```{r}
sound |> 
  filter(cycle == 0) |> 
  ggplot(
    aes(x, y, cycle = cycle)
  ) +
 geom_point(
    aes(
      nudge_x = after_stat(nudge) * 0.06
    ),
    stat = "manual",
    fun = make_cycle2,
    position = position_nudge()
  ) +
  geom_line(
    data = wave |> filter(cycle == 0),
    stat = "manual",
    fun = make_sine2,
    aes(
       y = (after_stat(y) * amp) - 0.21
    )
  ) +
  theme_no_x()+
  theme_no_y()
```

Where air is tightly packed, there's high pressure.
Where it's more spread out, low pressure.

## Sound

```{r}
#| label: fig-sound
sound |> 
  ggplot(
    aes(x,y, cycle = cycle)
  ) +
  geom_point(
    aes(
      nudge_x = after_stat(nudge) * 0.045
    ),
    stat = "manual",
    fun = make_cycle2,
    position = position_nudge()
  ) + 
  geom_line(
    data = wave,
    stat = "manual",
    fun = make_sine2,
    aes(
       y = (after_stat(y) * amp) - 0.2
    )
  ) +
  transition_time(cycle) +
  coord_fixed()+
  ease_aes('linear') +
  labs(title = 'cycle: {sprintf("%.2f", frame_time)}') +
  theme_no_x()+
  theme_no_y()
```

## Sounds

```{r}
#| fig-width: 6
#| fig-height: 2
tibble(
  x = seq(0, 2, length = 500),
  y = sin(x * 2 * 2 * pi)
) |> 
  data2shape(ob_point) |> 
  ob_path()->
  signal
  
tibble(
  x = seq(0, 2, length = 500),
  y = sin((x+0.125) * 2 * 2 * pi)
) |> 
  data2shape(ob_point) |> 
  ob_path() ->
  signal2

peaks <- ob_point(
  x = c(0.625, 0.625+ 0.5),
  y = c(1,1),
  color = ptol_red
)
```

```{r}
ggdiagram(linewidth = 1, font_size = 16) +
  {
    baseline <- ob_segment(
      ob_point(0, 0),
      ob_point(2, 0),
      color = "grey50",
      hjust = 0
    ) 
  } +
  signal +
  ob_segment(
    ob_point(-0.125, -1) ,
    ob_point(-0.125, 1),
    color = "grey50",
    label = ob_label("pressure", nudge_y = 0.3)
  ) +
  ob_label(
    "space",
    baseline@p1,
    color = "grey50"
  ) +  
  peaks +
  {
    peak1 <- peaks[1]
    peak1@x <- 0.125
    peak1
  } +
  ob_label(
    "peak",
    peak1,
    nudge_y = 0.08
  ) +
  {
    trough <- ob_point(
      0.125 + 0.25,
      -1,
      color = ptol_blue
    )
  } +
  ob_label(
    "trough",
    trough,
    nudge_y = -0.1
  ) +
  connect(
    peaks[1], 
    peaks[2],
    label = "wavelength",
    color = ptol_red
  ) +
  {
    peak3 <- ob_point(
      (0.5 * 3) + 0.125,
      1,
      color = ptol_green
    )
  } +
  {
    trough2 <- ob_point(
      ((0.5 * 3) + 0.125) - 0.25,
      -1,
      color = ptol_green
    )
  } +
  connect(
    ob_point(
      trough2@x,0
    ),
    trough2,
    label = ob_label("amplitude")
  ) +
  connect(
    ob_point(
      peak3@x,0
    ),
    peak3,
    label = ob_label("amplitude")
  )  +
  theme(
    aspect.ratio = 1/2
  )
```

## Sounds

```{r}
cols <- scico(5, palette = "glasgow")
phase_shifts <- seq(0, 0.33, length = 5)

imap(
  phase_shifts,
  \(p, idx){
    tibble(
      x = seq(0, 2, length = 200),
      y = sin((x-p) * 2 * 2 * pi)
    ) |> 
    data2shape(ob_point) |> 
    ob_path(
      color = cols[idx]
    )
  }
) -> 
  phases

imap(
  phase_shifts,
  \(p, idx){
    ob_point(
      0.125 + p,
      1
    )
  }
) -> 
  phase_peaks

ggdiagram(linewidth = 1, font_size = 16) + 
  phases +
  connect(
    ob_point(0.125, 1.1),
    ob_point(0.124 + 0.33, 1.1),
    label = ob_label("phase")
  ) +
  theme(
    aspect.ratio = 1/2
  )
```

## Sounds - Frequency

Frequency = how many high pressure zones per second

This is going to be a function of both the wavelength and the speed of the wave.

## Sounds - Frequency

```{r}
make_cycle3 <- function(df){
  df |> 
    mutate(
      nudge = cos((x - cycle) * freq * 2 * pi) * nscale
    )
}

make_sine3 <- function(df){
  df |> 
    mutate(
      y =  sin((x - cycle) * freq * 2 * pi)
    )
}
```

```{r}
#| fig-width: 5
#| fig-height: 3
wave |> 
  ggplot(
    aes(x, cycle = cycle, freq = 2)
  ) +
  geom_line(
    data = wave |>  mutate(speed = "fast"),
    stat = "manual",
    fun = make_sine3,
    aes(
       y = (after_stat(y) ) 
    )
  ) + 
  geom_line(
    data = wave |>  mutate(speed = "slow"),
    aes(
      cycle = cycle * 0.3,
      freq = 2,
      y = after_stat(y) 
    ),
    stat = "manual",
    fun = make_sine3
  ) +   
  transition_time(cycle) +
  ease_aes('linear') +
  labs(title = 'cycle: {sprintf("%.2f", frame_time)}') +
  theme_no_x()+
  theme_no_y() + 
  facet_wrap(~speed, ncol = 1)
```

## Sound - Frequency

Because the speed of sound will usually be fixed, Wavelength $\propto$ Frequency

$$
34000 \frac{cm}{s}
$$

## Sound - Frequency

Concert A is 440 Hz (that is, $\frac{1}{s}$)

$$
\frac{speed}{freq} = \frac{\frac{cm}{s}}{\frac{1}{s}}=\frac{cm}{s}\frac{s}{1} = cm
$$

```{r}
#| echo: true
34000/440
```

## Sound - Increasing Frequency

```{r}
sound |> 
  ggplot(
    aes(x,y, cycle = cycle, freq = 1.5 + (cycle/2), nscale = 0.03)
  ) +
  geom_point(
    aes(
      nudge_x = after_stat(nudge)
    ),
    stat = "manual",
    fun = make_cycle3,
    position = position_nudge()
  ) + 
  geom_line(
    data = wave,
    stat = "manual",
    fun = make_sine3,
    aes(
       y = (after_stat(y) * amp) - 0.2
    )
  ) +
  transition_time(cycle) +
  coord_fixed()+
  ease_aes('linear') +
  labs(title = 'cycle: {sprintf("%.2f", frame_time)}') +
  theme_no_x()+
  theme_no_y()
```

## Sound - Frequency

Every time frequency doubles, we experience it as the frequency increasing 1 octave.

<iframe src="https://strudel.cc/#ZnJlcSgiNTUiLm11bCgiPDEgMiA0IDggMTY%2BKjIiKSkKICAucygic2luZSIpCiAgLnNlZ21lbnQoNCkKICAuc2NvcGUoKQ%3D%3D" width="500" height="500">

</iframe>

## Sound - Intensity

A relationship between frequency *and* amplitude

```{r}
#| fig-width: 5
#| fig-height: 3
expand_grid(
  amplitude = seq(0.01, 0.05, length = 100),
  frequency = c(110, 220, 440)
) |> 
  mutate(
    intensity = 2 * (pi^2) * (amplitude^2) * (frequency^2)
  ) |> 
  ggplot(
    aes(amplitude, intensity)
  )+
  geom_textline(
    aes(
      color = factor(frequency),
      label = str_glue("{frequency} Hz")
    ),
    show.legend = F,
    linewidth = 1,
    size = 4
  ) +
  theme_sub_axis(
    text = element_blank()
  )
```

## Sound - Increasing Amplitude

```{r}
sound |> 
  ggplot(
    aes(x,y, cycle = cycle, freq = 1.5, nscale = 0.03 + (0.1 * (cycle/3)))
  ) +
  geom_point(
    aes(
      nudge_x = after_stat(nudge)
    ),
    stat = "manual",
    fun = make_cycle3,
    position = position_nudge()
  ) + 
  geom_line(
    data = wave,
    stat = "manual",
    fun = make_sine3,
    aes(
       y = (after_stat(y) * (after_stat(nscale)*2)) - 0.3
    )
  ) +
  transition_time(cycle) +
  coord_fixed()+
  ease_aes('linear') +
  labs(title = 'cycle: {sprintf("%.2f", frame_time)}') +
  theme_no_x()+
  theme_no_y()
```

## Sound - Loudness

Decibels (dB) are a measure of *relative* intensity.

$$
1 dB = 10\log_{10}\left(\frac{s_1}{s_2}\right)
$$

## Sound - Doppler Effect

```{r}
make_circle <- function(df){
  df |> 
    mutate(
      z = x
    ) |> 
    reframe(
      .by = z,
      theta = seq(0, 2*pi, length = 100),
      x = (cos(theta) * ((time-z)*1.2)) + z,
      y = sin(theta) * ((time-z)*1.2),
      across(
        matches("[^xyz]"),
        first
      )
    ) |> 
    filter(
      (time-z) >= 0
    )
}
```

```{r}
expand_grid(
  x = c(seq(0, 3, length = 10)),
  time = c(seq(0,3, length = 10)),
  dist1 = sqrt(((time-x)*1.2)^2)
) ->
  dop

tibble(
  x = seq(0, 3, length = 10),
  dist_to_y = sqrt((x^2) + (0.5^2)),
  time = (dist_to_y/1.2) + x
) |> 
  filter(time <= 3) ->
  pulse

dop |> 
  ggplot(
    aes(x, time = time)
  ) + 
  geom_point(
    aes(x = time, y = 0)
  )+
  geom_polygon(
    aes(color = x, group = x),
    stat = "manual",
    fun = make_circle,
    fill = NA,
    show.legend = F
  )+
  coord_fixed(ylim = c(-1,1), xlim = c(-0.2, 3))+
  transition_time(time)+
  theme_sub_axis(
    text = element_blank()
  )+
  labs(
    title = 'time: {sprintf("%.2f", ifelse(frame_time <= 3, frame_time, 3))}',
    x = "space",
    y = "space"
  )
```
